---
title: "Introduction to Data Science Project 1"
author: "Group 2"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: False
    toc: yes
    toc_depth: 3
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, results = 'markup', message = FALSE)
options(scientific=T, digits = 3) 
```

```{r ezids, include=F}
library(readr)
library(dplyr)
library(ggplot2)
library(forecast)
library(sp)
library(caret)

```

# Explore and Clean the Data

```{r, echo=TRUE}
data <- read.csv("Crime_Incidents_in_2024.csv")
str(data)
```
```{r, echo=TRUE}
head(data, 5)
```


# Question 1: Frequency of crime across neighborhoods

```{r, echo=TRUE}
neighborhood_crime_frequency <- data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(crime_frequency = n()) %>%
  arrange(desc(crime_frequency))
```


```{r, echo=TRUE}
# Visualize crime frequency using a bar chart
ggplot(neighborhood_crime_frequency, 
       aes(x = NEIGHBORHOOD_CLUSTER, 
           y = crime_frequency)) + 
  geom_bar(stat = "identity") + 
  theme_classic() + 
  labs(title = "Crime Frequency by Neighborhood", 
       x = "Neighborhood", 
       y = "Crime Frequency") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) + 
  scale_x_discrete(expand = c(0.0001, 0.0001))


```
```{r, echo=TRUE}
# 1. Crime frequency across different neighborhoods
crime_by_neighborhood <- data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(Total_Crimes = n(), .groups = "drop") %>%
  arrange(desc(Total_Crimes))

# Plot crime frequency by neighborhood
ggplot(crime_by_neighborhood, aes(x = reorder(NEIGHBORHOOD_CLUSTER, -Total_Crimes), y = Total_Crimes)) + 
  geom_bar(stat = "identity", fill = "darkblue") + 
  theme_classic() + 
  labs(title = "Crime Frequency Across Neighborhoods in DC - 2024", x = "Neighborhood Cluster", y = "Total Crimes") + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1, size = 8)) + 
  scale_x_discrete(expand = c(0.0001, 0.0001))


```


```{r, echo=TRUE}

# Step 3: Group by NEIGHBORHOOD_CLUSTER and count the number of crimes
crime_summary <- data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(crime_frequency = n())

# Step 4: Calculate the variance
variance <- var(crime_summary$crime_frequency)

# Step 5: Calculate the standard deviation
standard_deviation <- sd(crime_summary$crime_frequency)

# Step 6: Print the results
cat("Variance of Crime Frequencies:", variance, "\n")
cat("Standard Deviation of Crime Frequencies:", standard_deviation, "\n")

```
```{r, echo=TRUE}
library(e1071)
library(dplyr)

# Step 3: Summarize crime frequency by NEIGHBORHOOD_CLUSTER
crime_summary <- data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(crime_frequency = n())  # Count occurrences per neighborhood

# Step 4: Calculate skewness
crime_skewness <- skewness(crime_summary$crime_frequency)

# Step 5: Print the skewness value
cat("Skewness of Crime Frequency Distribution:", crime_skewness, "\n")

```

# Question 2: Time of year with highest and lowest crime rates

```{r, echo=TRUE}
monthly_crime_rate <- data %>%
  mutate(MONTH = format(as.Date(REPORT_DAT), "%m")) %>%
  group_by(MONTH) %>%
  summarise(crime_rate = n()) %>%
  arrange(desc(crime_rate))
```

```{r, echo=TRUE}
ggplot(monthly_crime_rate, aes(x = MONTH, y = crime_rate)) + 
  geom_point() + 
  theme_classic() + 
  labs(title = "Crime Rate by Month", x = "Month", y = "Crime Rate")

```
# Question 3: Most common types of crime reported


```{r, echo=TRUE}
crime_type_frequency <- data %>%
  group_by(OFFENSE) %>%
  summarise(frequency = n()) %>%
  arrange(desc(frequency))

```

```{r, echo=TRUE}
ggplot(crime_type_frequency, aes(x = OFFENSE, y = frequency)) + 
  geom_bar(stat = "identity") + 
  theme_classic() + 
  labs(title = "Crime Type Frequency", x = "Crime Type", y = "Frequency") + 
  scale_x_discrete(expand = c(0.2, 0.2)) + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))


```
# Question 4: Correlation between time of day and crime type


```{r, echo=TRUE}

data$HOUR <- format(as.Date(data$REPORT_DAT), "%H")
hourly_crime_type_frequency <- data %>%
  group_by(HOUR, OFFENSE) %>%
  summarise(frequency = n(), .groups = "drop")

```

```{r, echo=TRUE}

library(pheatmap)
pheatmap(hourly_crime_type_frequency_cast, 
         main = "Correlation between Time of Day and Crime Type", 
         cluster_rows = FALSE, 
         cluster_cols = FALSE)


```
# Question 5: Neighborhoods with highest rates of homicide

```{r, echo=TRUE}
homicide_data <- data %>%
  filter(OFFENSE == "Homicide")
neighborhood_homicide_frequency <- homicide_data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(homicide_frequency = n()) %>%
  arrange(desc(homicide_frequency))

```

```{r, echo=TRUE}


# Step 2: Filter for homicide offenses
homicide_data <- data %>%
  filter(OFFENSE == "HOMICIDE")

# Step 3: Summarize homicides by neighborhood
neighborhood_homicide_summary <- homicide_data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(homicide_count = n()) %>%
  arrange(desc(homicide_count))  # Sort in descending order

# Step 4: Print the top neighborhoods with the highest homicide counts
print(head(neighborhood_homicide_summary))  # Display the top neighborhoods

# Step 5 (Optional): Visualize homicide counts by neighborhood
ggplot(neighborhood_homicide_summary, aes(x = reorder(NEIGHBORHOOD_CLUSTER, -homicide_count), 
                                          y = homicide_count, fill = NEIGHBORHOOD_CLUSTER)) +
  geom_col() +
  theme_classic() +
  labs(title = "Homicide Counts by Neighborhood in DC", 
       x = "Neighborhood Cluster", 
       y = "Homicide Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_brewer(palette = "Set3")


```

# Analyze the spatial distribution using latitude and longitude:

```{r, echo=TRUE}
# Check the range of latitude and longitude
range(data$LATITUDE, na.rm = TRUE)
range(data$LONGITUDE, na.rm = TRUE)

# Optional: Use ggplot2 for geographic visualization if you have map boundaries
ggplot(data, aes(x = LONGITUDE, y = LATITUDE)) +
  geom_point(alpha = 0.5) +
  theme_minimal() +
  labs(title = "Geographic Distribution of Incidents", x = "Longitude", y = "Latitude")

```



# Identifying missing values and Outliers

```{r, echo=TRUE}
# Missing values in LATITUDE and LONGITUDE
missing_geolocation <- sum(is.na(data$LATITUDE) | is.na(data$LONGITUDE))
print(paste("Missing geolocation data:", missing_geolocation))

# Identify outliers in LATITUDE and LONGITUDE
summary(data$LATITUDE)
summary(data$LONGITUDE)

```
- The latitude values range from 38.8 to 39.0, with a median and mean value of around 38.9 while the longitude values range from  -77.1 to -77.0, with a median and mean value of around -77.0.


# T-test Comparing mean latitude between two shifts (Midnight and Day)
```{r, echo=TRUE}

t_test_data <- data %>%
  filter(SHIFT %in% c("MIDNIGHT", "DAY"))  

t_test_result <- t.test(LATITUDE ~ SHIFT, data = t_test_data)  
print(t_test_result)


```
The results show that there is no significant difference as the p-value (0.1) is greater than the typical significance level (0.05) showing that there is no statistically significant difference in the means of LATITUDE between the DAY and MIDNIGHT groups. Furthermore, the 95% confidence interval (-0.06121, 0.00866) overlaps with 0 and the sample estimates show that the mean LATITUDE values for both DAY and MIDNIGHT groups are equal (38.9)

# Anova Comparing more than two groups : Comparing mean latitude across neighborhoods

```{r, echo=TRUE}
anova_result <- aov(LATITUDE ~ NEIGHBORHOOD_CLUSTER, data = data)
summary(anova_result)
```

# Chi-square evaluating relationship between two categorical variables
```{r, echo=TRUE}
# Create a contingency table
contingency_table <- table(data$OFFENSE, data$SHIFT)

# Perform the chi-square test
chi_square_result <- chisq.test(contingency_table)
print(chi_square_result)

```