---
title: "Introduction to Data Science Project 2"
author: "Group 2"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: show
    number_sections: False
    toc: yes
    toc_depth: 3
    toc_float: yes
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(warning = FALSE, results = 'markup', message = FALSE)
options(scientific=T, digits = 3) 
```

```{r ezids, include=F}

library(tidyverse)
library('ezids')

```

# Loading the Dataset and Handling missing values

```{r, echo=TRUE}

crime_data <- read.csv("Crime_Incidents_in_2024.csv")
```

```{r, echo=TRUE}

nrow(crime_data)
```

```{r, echo=TRUE}

colSums(is.na(crime_data))
```
```{r, echo=TRUE}

missing_crime_data <- crime_data == "" | is.na(crime_data)
colSums(missing_crime_data)
```

```{r, echo=TRUE}


crime_data_clean <- crime_data[!apply(is.na(crime_data), 1, all), ]

crime_data_clean$WARD[is.na(crime_data_clean$WARD)] <- median(crime_data_clean$WARD, na.rm = TRUE)
crime_data_clean$DISTRICT[is.na(crime_data_clean$DISTRICT)] <- "Unknown"
crime_data_clean$CENSUS_TRACT[is.na(crime_data_clean$CENSUS_TRACT)] <- median(crime_data_clean$CENSUS_TRACT, na.rm = TRUE)
```

```{r, echo=TRUE}
colSums(is.na(crime_data_clean))
```
# Basic Statistics 
   
```{r, echo=TRUE}
crime_data_clean <- crime_data_clean %>% select(-OCTO_RECORD_ID)
crime_data_clean$PSA[is.na(crime_data_clean$PSA)] <- median(crime_data_clean$PSA, na.rm = TRUE)
```

```{r, echo=TRUE}
colSums(is.na(crime_data_clean))
```


```{r, echo=TRUE}
summary(crime_data_clean$LATITUDE)
summary(crime_data_clean$LONGITUDE)
summary(crime_data_clean$WARD)
summary(crime_data_clean$PSA)
```

# Exploratory Data Analysis

```{r, echo=TRUE}
library(ggplot2)

library(dplyr)

top_offenses <- crime_data_clean %>%
  count(OFFENSE, sort = TRUE) %>%
  top_n(10)

ggplot(top_offenses, aes(x = reorder(OFFENSE, n), y = n)) +
  geom_bar(stat = "identity", fill = "steelblue") +
  coord_flip() +  # Horizontal bars for better readability
  theme_minimal() +
  labs(title = "Top 10 Most Common Crimes", x = "Offense Type", y = "Count")

```

```{r, echo=TRUE}
ggplot(crime_data_clean, aes(x = SHIFT, fill = OFFENSE)) +
  geom_bar(position = "dodge") +
  theme_minimal() +
  labs(title = "Crime Distribution by Police Shift", x = "Police Shift", y = "Count")

```
# Statistics 

```{r, echo=TRUE}

crime_data_clean$Is_Felony <- ifelse(crime_data_clean$OFFENSE %in% c("HOMICIDE", "ROBBERY", "BURGLARY", "ASSAULT W/DANGEROUS WEAPON", "SEX ABUSE", "ARSON"), 1, 0)

```

```{r, echo=TRUE}

table(crime_data_clean$Is_Felony)

```
```{r, echo=TRUE}

summary(crime_data_clean)  

crime_data_clean %>%
  summarise(
    Total_Felonies = sum(Is_Felony), 
    Mean_Felonies_Per_Record = mean(Is_Felony), 
    SD_Felonies = sd(Is_Felony), 
    Median_Felonies = median(Is_Felony) 
  )
```

# Correlation Between Felony Counts & Police Districts/Wards

```{r, echo=TRUE}

library(dplyr)

felony_by_district <- crime_data_clean %>%
  group_by(DISTRICT) %>%
  summarise(Felony_Count = sum(Is_Felony))

felony_by_ward <- crime_data_clean %>%
  group_by(WARD) %>%
  summarise(Felony_Count = sum(Is_Felony))
```

```{r, echo=TRUE}

crime_data_clean$DISTRICT <- as.numeric(as.factor(crime_data_clean$DISTRICT))
crime_data_clean$WARD <- as.numeric(crime_data_clean$WARD)
```

```{r, echo=TRUE}
cor.test(crime_data_clean$DISTRICT, crime_data_clean$Is_Felony)
cor.test(crime_data_clean$WARD, crime_data_clean$Is_Felony)
```


# Linear Regression - Predicting Felony Counts

```{r, echo=TRUE}
lm_model <- lm(Is_Felony ~ WARD + DISTRICT + METHOD + BID, data = crime_data_clean)
summary(lm_model)
```

# Logistic Regression: Does Felony Offense Occur More During Certain Police Shifts?

```{r, echo=TRUE}

log_model <- glm(Is_Felony ~ METHOD + SHIFT + NEIGHBORHOOD_CLUSTER, data = crime_data_clean, family = binomial)
summary(log_model)
```

## Model Performance Check

```{r, echo=TRUE}

predicted_probs <- predict(log_model, type = "response")
```

```{r, echo=TRUE}

predicted_felonies <- ifelse(predicted_probs > 0.5, 1, 0)
```

```{r, echo=TRUE}

library(caret)
confusionMatrix(as.factor(predicted_felonies), as.factor(crime_data_clean$Is_Felony))
```

# ANOVA Test on Crime Method : Does Crime Method Influence Felony Classification?

```{r, echo=TRUE}

anova(lm(Is_Felony ~ METHOD, data = crime_data_clean))
```
# Tukey’s Post-Hoc Test

```{r, echo=TRUE}

library(multcomp)
crime_data_clean$METHOD <- as.factor(crime_data_clean$METHOD)
tukey_test <- glht(lm(Is_Felony ~ METHOD, data = crime_data_clean), linfct = mcp(METHOD = "Tukey"))
summary(tukey_test)

```


```{r, echo=TRUE}

library(ggplot2)
ggplot(crime_data_clean, aes(x = METHOD, fill = as.factor(Is_Felony))) +
  geom_bar(position = "fill") +
  labs(title = "Felony Rates by Crime Method", x = "Crime Method", y = "Proportion of Felonies") +
  theme_minimal()
```

# T-Test for Felony Rate Differences

```{r, echo=TRUE}

crime_data_clean$BID_presence <- ifelse(crime_data_clean$BID != "", "BID", "Non-BID")
```


```{r, echo=TRUE}

table(crime_data_clean$BID_presence)
```
```{r, echo=TRUE}

t.test(Is_Felony ~ BID_presence, data = crime_data_clean, var.equal = TRUE)
```

#  Felony Crime Heatmap

```{r, echo=TRUE}

library(ggplot2)

ggplot(crime_data_clean, aes(x = LONGITUDE, y = LATITUDE, color = as.factor(Is_Felony))) +
  geom_point(alpha = 0.5) +
  scale_color_manual(values = c("blue", "red")) +
  labs(title = "Felony vs. Non-Felony Crime Hotspots",
       x = "Longitude", y = "Latitude", color = "Crime Type") +
  theme_minimal()
```
```{r, echo=TRUE}

ggplot(crime_data_clean, aes(x = BID_presence, fill = as.factor(Is_Felony))) +
  geom_bar(position = "fill") +
  labs(title = "Felony Rates in BID vs. Non-BID Areas",
       x = "BID Area", y = "Proportion of Felonies") +
  theme_minimal()
```
```{r, echo=TRUE}

ggplot(crime_data_clean, aes(x = DISTRICT, fill = as.factor(Is_Felony))) +
  geom_bar() +
  labs(title = "Felony Crime Counts Across Police Districts",
       x = "Police District", y = "Number of Crimes") +
  theme_minimal()
```

# Classification & Regression Trees(CART)

```{r, echo=TRUE}

library(rpart)
library(rpart.plot)

tree_model <- rpart(Is_Felony ~ WARD + DISTRICT + METHOD + SHIFT + BID_presence,
                    data = crime_data_clean, method = "class")

rpart.plot(tree_model, type = 3, box.palette = "RdBu", fallen.leaves = TRUE)
```

```{r, echo=TRUE}

predicted_tree <- predict(tree_model, type = "class")
```

```{r, echo=TRUE}

library(caret)
confusionMatrix(as.factor(predicted_tree), as.factor(crime_data_clean$Is_Felony))
```

```{r, echo=TRUE}

library(caret)
varImp(tree_model)
```


# Cross-validation for decision tree reliability


```{r, echo=TRUE}
library(caret)
train_control <- trainControl(method = "cv", number = 10)
tree_model_cv <- train(Is_Felony ~ WARD + DISTRICT + METHOD + SHIFT + BID_presence, 
                       data = crime_data_clean, 
                       method = "rpart", 
                       trControl = train_control)
print(tree_model_cv)
```
# Summary of Key Findings
```{r, echo=TRUE}

crime_data_clean$Is_Felony <- as.numeric(as.character(crime_data_clean$Is_Felony))

crime_summary <- crime_data_clean %>%
  summarise(
    Total_Felonies = sum(Is_Felony), 
    Mean_Felonies = mean(Is_Felony), 
    SD_Felonies = sd(Is_Felony), 
    Median_Felonies = median(Is_Felony)
  )

print(crime_summary)

```

# Final Crime Trend Visualization
```{r, echo=TRUE}

ggplot(crime_data_clean, aes(x = METHOD, fill = as.factor(Is_Felony))) +
  geom_bar(position = "fill") +
  labs(title = "Felony Rates by Crime Method", x = "Crime Method", y = "Proportion of Felonies") +
  theme_minimal()
```
