<<<<<<< HEAD
---
title: "Project_1"
author: "Sadheen Hossain"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

library(ezids)
library(tidyverse)
library(ggplot2)
install.packages("magrittr")
library(magrittr)

```{r}
crime_data <- read.csv("Crime_Incidents_in_2024.csv", stringsAsFactors = FALSE)
head(crime_data)
```
```{r}
# 1. Crime frequency across different neighborhoods
crime_by_neighborhood <- crime_data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(Total_Crimes = n(), .groups = "drop") %>%
  arrange(desc(Total_Crimes))

# Plot crime frequency by neighborhood
ggplot(crime_by_neighborhood, aes(x=reorder(NEIGHBORHOOD_CLUSTER, -Total_Crimes), y=Total_Crimes)) + 
  geom_bar(stat="identity", fill="darkred") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Crime Frequency Across Neighborhoods in DC - 2024")
```
```{r}
# 2. Crime rates by time of year

colnames(crime_data)[colnames(crime_data) %in% c("REPORT_DAT ", "report_dat", "Report Date")] <- "REPORT_DAT"
crime_data <- crime_data %>% filter(!is.na(REPORT_DAT))

# Number of crimes per month
crime_by_month <- crime_data %>%
  group_by(Month) %>%
  summarise(Total_Crimes = n()) %>%
  arrange(Month)

# Plot crime rates by month
ggplot(crime_by_month, aes(x=Month, y=Total_Crimes)) + 
  geom_bar(stat="identity", fill="steelblue") + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Crime Rates by Month in 2024")
```
```{r}
crime_types <- crime_data %>%
  group_by(OFFENSE) %>%
  summarise(Total_Crimes = n()) %>%
  arrange(desc(Total_Crimes))

ggplot(crime_types, aes(x=reorder(OFFENSE, -Total_Crimes), y=Total_Crimes)) + 
  geom_bar(stat="identity", fill="red") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Most Common Crimes in 2024")
```
```{r}
# 4. Correlation between time of day and crime type
crime_by_shift <- crime_data %>%
  group_by(SHIFT, OFFENSE) %>%
  summarise(Total_Crimes = n(), .groups = "drop") %>%
  arrange(desc(Total_Crimes))

ggplot(crime_by_shift, aes(x=SHIFT, y=Total_Crimes, fill=OFFENSE)) + 
  geom_bar(stat="identity", position="dodge") + 
  ggtitle("Crime Type by Time of Day") +
  theme_minimal()
```

```{r}
# 5. Neighborhood clusters with highest homicide rates
homicide_by_neighborhood <- crime_data %>%
  filter(OFFENSE == "HOMICIDE") %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(Total_Homicides = n()) %>%
  arrange(desc(Total_Homicides))

ggplot(homicide_by_neighborhood, aes(x=reorder(NEIGHBORHOOD_CLUSTER, -Total_Homicides), y=Total_Homicides)) + 
  geom_bar(stat="identity", fill="darkred") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Neighborhood Clusters with Highest Homicide Rates")
=======
---
title: "Project_1"
author: "Sadheen Hossain"
date: "`r Sys.Date()`"
output:
  html_document:
    code_folding: hide
    number_sections: false
    toc: yes
    toc_depth: 3
    toc_float: yes
  pdf_document:
    toc: yes
    toc_depth: '3'
---

library(ezids)
library(tidyverse)
library(ggplot2)
install.packages("magrittr")
library(magrittr)

```{r}
crime_data <- read.csv("Crime_Incidents_in_2024.csv", stringsAsFactors = FALSE)
head(crime_data)
```
```{r}
# 1. Crime frequency across different neighborhoods
crime_by_neighborhood <- crime_data %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(Total_Crimes = n(), .groups = "drop") %>%
  arrange(desc(Total_Crimes))

# Plot crime frequency by neighborhood
ggplot(crime_by_neighborhood, aes(x=reorder(NEIGHBORHOOD_CLUSTER, -Total_Crimes), y=Total_Crimes)) + 
  geom_bar(stat="identity", fill="darkred") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Crime Frequency Across Neighborhoods in DC - 2024")
```
```{r}
# 2. Crime rates by time of year

colnames(crime_data)[colnames(crime_data) %in% c("REPORT_DAT ", "report_dat", "Report Date")] <- "REPORT_DAT"
crime_data <- crime_data %>% filter(!is.na(REPORT_DAT))

# Number of crimes per month
crime_by_month <- crime_data %>%
  group_by(Month) %>%
  summarise(Total_Crimes = n()) %>%
  arrange(Month)

# Plot crime rates by month
ggplot(crime_by_month, aes(x=Month, y=Total_Crimes)) + 
  geom_bar(stat="identity", fill="steelblue") + 
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_minimal() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + 
  ggtitle("Crime Rates by Month in 2024")
```
```{r}
crime_types <- crime_data %>%
  group_by(OFFENSE) %>%
  summarise(Total_Crimes = n()) %>%
  arrange(desc(Total_Crimes))

ggplot(crime_types, aes(x=reorder(OFFENSE, -Total_Crimes), y=Total_Crimes)) + 
  geom_bar(stat="identity", fill="red") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Most Common Crimes in 2024")
```
```{r}
# 4. Correlation between time of day and crime type
crime_by_shift <- crime_data %>%
  group_by(SHIFT, OFFENSE) %>%
  summarise(Total_Crimes = n(), .groups = "drop") %>%
  arrange(desc(Total_Crimes))

ggplot(crime_by_shift, aes(x=SHIFT, y=Total_Crimes, fill=OFFENSE)) + 
  geom_bar(stat="identity", position="dodge") + 
  ggtitle("Crime Type by Time of Day") +
  theme_minimal()
```

```{r}
# 5. Neighborhood clusters with highest homicide rates
homicide_by_neighborhood <- crime_data %>%
  filter(OFFENSE == "HOMICIDE") %>%
  group_by(NEIGHBORHOOD_CLUSTER) %>%
  summarise(Total_Homicides = n()) %>%
  arrange(desc(Total_Homicides))

ggplot(homicide_by_neighborhood, aes(x=reorder(NEIGHBORHOOD_CLUSTER, -Total_Homicides), y=Total_Homicides)) + 
  geom_bar(stat="identity", fill="darkred") + 
  theme(axis.text.x = element_text(angle=45, hjust=1)) + 
  ggtitle("Neighborhood Clusters with Highest Homicide Rates")
>>>>>>> f7ff5fcc2a1b584f583676045a26f2080370a436
```